Document Indexing Using Aho-Corasick State Machines

### 2011-10-26 report/article.tex
Paper down to 11 pages - sending for review
Naive comparison almost done

%----------------------------------------
% Abstract
%----------------------------------------
\begin{abstract}
The Aho-Corasick algorithm was originally proposed as a bibliographic
search mechanism, efficient enough to preclude the construction and
maintenance of a search index. As the size of document collections
have grown since the development of the algorithm, document
collections have grown to the point where cost of rescanning a corpus
for every search has become prohibitive and indexing, in some form or
another, is an essential optimization for any modern information
retrieval system. In this paper we demonstrate the use of
Aho-Corasick, not as an alternative to indexing but as a lexical
scanning tool employed in the construction of an inverted
index. Consistent with analytical predictions, an empirical evaluation
of our hybrid indexer, shows significant improvement over a
na\"{\i} implementation. In addition, the modular architecture of our indexer
allowed the indexing task to be parallelized over multiple physical
nodes. The results of parallelization are presented and offer
interesting insights towards further horizontal scaling.
\\\\
{\bf Keywords:} inverted index, Aho-Corasick state machine, finite
automata, pattern matching, parsing
\end{abstract}

@@

In order to construct an inverted index, documents must be scanned to
determine term frequencies. The Aho-Corasick string matching
algorithm\cite{RefWorks:103} is a simple and efficient text scanning algorithm. The
algorithm constructs a finite state machine to scan for a given set of
keywords. It is, in effect, a reduced grammar regular expression
parser of the type described in \cite{RefWorks:111}. The algorithm is simple and
efficient, construction time being proportional to the sum of the
length of the keyword set, and the number of state transitions
required to scan a document is independent of the number of the size
of the keyword set. Aho-Corasick was originally proposed as a
bibliographic search mechanism, efficient enough to preclude the
construction and maintenance of a search index. As the size of
document collections have grown since the publishing of the
Aho-Corasick algorithm, document indexing is essential for efficient
document search. In this paper we demonstrate the use of Aho-Corasick,
not as an alternative to indexing but as a lexical scanning tool
employed in the construction of an inverted index. We provide analytic
predictions of our implementation against a naive algorithm and
finally, an empirical evaluation of our inverted index implementation
is conducted and the results analyzed.

%----------------------------------------
% Design
%----------------------------------------
\section{Implementation}
\label{sec:implementation}
As can be seen in figure \ref{fig:deploymentmodel}, the indexer
processes are symmetrical and can deployed on multiple physical nodes. The
individual indexers do not interact directly with one another, making
for a simple deployment and operation model. The execution loop of
each indexer is as follows:

\begin{enumerate}
\item Initialize the indexer by retrieving a collection of index
  keywords and synonyms from a relational database. These keywords are
  used to construct a lexical parser which will be used to scan and
  index documents.
\item Retrieve a batch of unprocessed documents from a relational
  database. The document batch will be sized according to the physical
  capabilities of each node. In this implementation, this tuning task
  is a manual exercise but future enhancements may include an adaptive
  loading component.
\item Parse each document retrieved and construct an inverted index
  representing the batch. This \textit{delta} index, as we shall call
  it, is then used to augment the global index maintained in a
  relational database.
\item Repeat from step 2.
\end{enumerate}

As shown in figure \ref{fig:deploymentmodel}, the searcher component
can be executed from any node which has access to the relational
database housing the document index. The execution path of a single
search query wold be as follows:

\begin{enumerate}
  \item Canonize search terms based on keyword synonyms defined in the
    keyword store.
  \item Execute a boolean query against the inverted index.
  \item Return the list of corpus documents containing the
    intersection of the canons of the search terms.
\end{enumerate}

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.70\textwidth,height=!]{deploymentmodel}
  \end{center}
  \caption{Indexer deployment model}
  \label{fig:deploymentmodel}
\end{wrapfigure}

%----------------------------------------
% Experiment Results
%----------------------------------------
\section{Experiment Results}
\label{sec:experimentresults}
The primary goal of this project is to be able to reduce the time
to build a document index. As our system is building the inverted index by
parsing all documents of the collection, we can expect the
time complexity of the task to be proportional to the sum total of
characters in the collection. As a result, changing the  
number of documents or the average size of each document should
have a big impact in our time complexity. In addition, our
implementation will incur the cost of building the Aho-Corasick state
machines used to scan the documents. We can expect the time to build a
state machine to be proportional to the sum of the size of the
keywords.

Finally we will do a time comparison between our indexer
implementation and a na\"{i}ve implementation. All the time
complexities tests are done on a single computer and it is important
to understand that all these processes can be parallelized. In section
\ref{parallelization} we install our indexer on multiple computers and
explore the performance gain of running parallel instances
simultaneously.

All tests were run on a single Intel Core 2 Duo 2GHz processor, 1GB
RAM, running a 64-bit Linux 2.6.35 SMP kernel. The Java Virtual
Machine used was version 1.6.0-24. The database used was MySQL,
running on the same node as the indexer.

%----------------------------------------
% Number of Keywords
%----------------------------------------
\subsection{Number of Keywords}
To be able to study the impact of different number of keywords, we
built a keywords database with 10,000 most popular English
words \cite{wordlist}. All the experiments in this
part are done with 50,000 posts randomly downloaded from the Internet. The
idea is to get a good distribution of documents with respect to length
and content. The experiment shown here consist of building the index
for different numbers of keywords from the 10,000 keyword database and
explore the time and space complexity for the building of the state
machine and the building of the index using the state machine.

The time taken to construct the Aho-Corasick state machine was
measured for different numbers of keywords. The results of this test
is shown in figure \ref{fig:numkeywordstimecomplexbuildsm} and
construction time can easily be considered linear with respect to the
number of keywords. This is consistent with \cite{RefWorks:103} which
proves that the state machine construction algorithm is linearly
proportional to the sum of the lengths of the keywords used to
construct the state machine.

The time taken for a state machines to build an inverted index of our
corpus was measured. The test was repeated using state machines
constructed with a varying number of keywords and the results plotted
in figure \ref{fig:numkeywordstimecomplexbuildindex}. According to
\cite{RefWorks:103} which proves that the number of state transitions
involved in processing an input string is independent of the number of
keywords used to construct the state machine, we would not expect the
time complexity to increase with the size of the keyword set. However,
as can be seen in \ref{fig:numkeywordstimecomplexbuildindex}, the time
to build the index increases logarithmically with the number of
keywords. This is likely due to the fact that as the number of
keywords increases, so does the size of the index and the number of
database interacts required to store and update the index.

@@

The space used by the index in the database is proportional to the
number of words we are using. It also means that our words have the
same probability at the beginning or at the end of our 12,000 word
dictionary.

%----------------------------------------
% Size of Corpus
%----------------------------------------
\subsection{Size of Corpus}
To be able to study the impact of different number of Posts, we built
a posts database with 400,000 posts randomly downloaded from
the Internet. Then the keywords used to build the index are simply the
first 5,000 from the previous keywords database.

The experiment shown here consist on building the index for different
number of Posts and explore the time and space complexity for the
building of the state machine and the building of the index once the
state machine is built. We can expect that the time to build the State
machine won’t change with the number of posts and the time to build
the index will increase almost linearly with the number of posts.

As we were expecting, the time to build the index with the state
machine strategy is linear with respect to the number of post, as
depicted in figure \ref{fig:corpsizetimecomplexbuildindex}. However, to be able
to perform this experiment we had to run the indexer program with
batches of 10,000 posts due to Java heap space problems. As a result, the state
machine is rebuilt every 10,000 posts.

The space used to store the index in the database is
proportional to the size of the corpus as shown in figure
\ref{fig:corpsizespacecomplexbuildindex}.

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{corpsizetimecomplexbuildindex}
    \caption{Time to build the inverted index as a function of the
      size of the corpus. Time taken is shown in milliseconds on the
      y-axis and the number of documents in the collection on the
      x-axis} 
    \label{fig:corpsizetimecomplexbuildindex}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{corpsizespacecomplexbuildindex}
    \caption{Size of index as a function of the size of the
      corpus. The number of database rows used to store the index is
      shown on the y-axis and the number of documents in the
      collection on the x-axis} 
    \label{fig:corpsizespacecomplexbuildindex}
  \end{minipage}
\end{figure}

%----------------------------------------
% Document Size
%----------------------------------------
\subsection{Document Size} 
As we have seen before, the time to build the index is almost
proportional with the number of post. Theoretically this time should
be proportional to the total length of all posts. Let’s focus on this
aspect and study the impact of the length of the post on the time to
build the state machine. To do that we built a 5,000 posts database
with posts that are all composed of 20,000 characters. Then, we ran
our system on different length of substring of these posts.

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{docsizetimecomplexbuildsm}
    \caption{Time to build Aho-Corasick state machine as a function of
      the average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis}
    \label{fig:docsizetimecomplexbuildsm}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{docsizetimecomplexbuildindex}
    \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis} 
    \label{fig:docsizetimecomplexbuildindex}
  \end{minipage}
\end{figure}

State machine construction time does not depend on the
length of the documents. Nonetheless it is interesting to compare
figures \ref{fig:docsizetimecomplexbuildsm} and
\ref{fig:docsizetimecomplexbuildindex}, showing the construction times
for the state machine and inverted index relative to the average
document size. The comparison shows that state machine construction
time is several orders of magnitude smaller than index construction
and negligible by comparison. Our state machine is built in 770 ms on
average for 1000 keywords with a standard deviation of 18ms.

As expected the time to build the indexer is almost linear with
respect to the average document size. This relationship is shown in
figure \ref{fig:docsizetimecomplexbuildindex}.

As expected, the larger the documents the more records created to
represent the index. However, the relationship is not linear since if
a word occurs many times in the same post we just save it once in the
database. This relationship is shown in figure
\ref{fig:docsizespacecomplexbuildindex}.

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.70\textwidth,height=!]{docsizespacecomplexbuildindex}
  \end{center}
  \caption{Size of index as a function of the average document
    size. The number of database rows used to store the index is
    shown on the y-axis and the average document size in characters
    on the x-axis}  
  \label{fig:docsizespacecomplexbuildindex}
\end{wrapfigure}

%----------------------------------------
% Comparison with Naive Index Builder
%----------------------------------------
\subsection{Comparison with Na\"{i}ve Index Builder}
This part focuses on the improvement that our system with respect to
time complexity in comparison with a na\"{i}ve index builder. The na\"{i}ve
indexer builder has the same input as our system i.e. a list of
keywords and a list of documents, and the same output i.e. an inverted
index corresponding to the keywords and documents. Algorithm
\ref{alg:invertedindex} is used in principal for both implementations.

@@

As we can see this algorithm has a complexity that directly depends on
the number of keywords $n$, the number of document $m$, and the length
of document since we have to check in each document for each
keyword. Moreover, we decided to limit, as much as possible, the
accesses to the database since it is really time consuming in
Java. That’s why we do one access to the database for each keywords
and not for each document. In the state machine we do as much accesses
to the database as the number of document (in the worst case meaning
we find at least one keyword per document). In the following
subsections we compare the results for this algorithm with our system
results for different number of keywords and posts.

For the na\"{i}ve algorithm described above, the theoretical worst
case time complexity relationship is shown in equation
\ref{eqn:timecomplexworstcase}.

\begin{equation}
\label{eqn:timecomplexworstcase}
Time \propto keyNum * ( docNum * maxLength + conn)
\end{equation}

where \(keyNum\) is the number of keywords, \(docNum\) is the number
of documents, \(maxLength\) is the maximum document length, and
\(conn\) is the time taken to create a connection to the database. For
the state machine algorithm the theoretical complexity relationship is
given by equation \ref{eqn:timecomplexsm}.

\begin{equation}
\label{eqn:timecomplexsm}
Time \propto docNum * ( maxLength + conn)
\end{equation}

To exmaine the effect of document size, the number of keywords was
fixed at 1,000 and the number of documents to 5,000 and we ran our
code on a database composed of document of different length. The
experiment results are shown in figure \ref{fig:naivedocumentsize}. As
expected, the curves for both our indexer and the naive indexer are
almost linear since for the naive implementation:

\[Time \propto (keyNum * docNum) * maxLength + conn * keyNum\]

and for our Aho-Corasick state machine method:

\[Time \propto docNum * maxLength + conn * docNum\]

We can observe that for a small length of document naive method is
faster since \(keyNum\) is smaller than \(docNum\). For \(docLength = 1300\)
our indexer is far more efficient. To see this effect more clearly,
experiment results are plotted on a logarithmic scale in figure
\ref{fig:naivedocumentsizelog}.As expected, the two implementations
exhibit a linear response to the number of keywords used. In the
na\"{i}ve implemementation:

@@

\[Time \propto (keyNum * maxLength) * docNum + conn * keyNum\]

\[Time \propto (maxLength + conn) * docNum \]

@@

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.60\textwidth,height=!]{parallelization}
  \end{center}
  \caption{Performance characteristics of parallel indexer
    deployment. The \textit{ideal} curve shows the theoretical
    performance curve based on linear combination of average
    individual node performance over the corpus} 
\end{wrapfigure}

>>>>

%----------------------------------------
% Abstract
%----------------------------------------
\begin{abstract}
The Aho-Corasick algorithm was originally proposed as a bibliographic
search mechanism, efficient enough to preclude the construction and
maintenance of a search index. As the size of document collections
have grown since the development of the algorithm, document
collections have grown to the point where cost of rescanning a corpus
for every search has become prohibitive and indexing, in some form or
another, is an essential optimization for any modern information
retrieval system. In this paper we demonstrate the use of
Aho-Corasick, not as an alternative to indexing but as a lexical
scanning tool employed in the construction of an inverted
index. Consistent with our analysis, an empirical evaluation of our
indexer implementation shows significant improvement over a na\"{\i}ve
algorithm. The modular architecture of our indexer allows the indexing
task to be parallelized over multiple physical nodes. The results of
parallelization are presented and offer interesting insights towards
further horizontal scaling.
\\\\
{\bf Keywords:} inverted index, Aho-Corasick state machine, finite
automata, pattern matching, parsing
\end{abstract}

@@

In order to construct an inverted index, documents must be scanned to
determine term frequencies. The Aho-Corasick string matching
algorithm\cite{RefWorks:103} is a simple and efficient text scanning
algorithm. The algorithm constructs a finite state machine to scan for
a given set of keywords. It is, in effect, a reduced grammar regular
expression parser of the type described in \cite{RefWorks:111}. The
algorithm is simple and efficient, construction time being
proportional to the sum of the length of the keyword set, and the
number of state transitions required to scan a document is independent
of the number of the size of the keyword set. In this paper we
demonstrate the use of Aho-Corasick, not as an alternative to indexing
but as a lexical scanning tool employed in the construction of an
inverted index. We provide analytic predictions of our implementation
against a na\"{\i}ve algorithm and finally, an empirical evaluation of
our inverted indexer implementation is conducted and the results
analyzed.

%----------------------------------------
% Design
%----------------------------------------
\section{Implementation}
\label{sec:implementation}
The Aho-Corasick indexer was implemented using Java and MySQL
relational database. Upon startup, the indexer reads a list of index
keywords and synonyms from the database and constructs an Aho-Corasick
state machine. The indexer then enters a loop in which it uses the
state machine to scan batches of documents retrieved from the
database. The document batch size is varied manually to match the
physical capabilites of the host. A possible enhancement to our
implementation could include an adaptive loading component. The output
of the state machine is used to augment a global index maintained in
the database.

Figure \ref{fig:deploymentmodel} shows how multiple indexer instances
can be deployed in parallel, using the database to synchronize their
access to the corpus and index. The individual indexers do not
interact directly with one another, making for a simple deployment and
operation model.

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.70\textwidth,height=!]{deploymentmodel}
  \end{center}
  \caption{Indexer deployment model showing multiple indexer instances
  operating in parallel}
  \label{fig:deploymentmodel}
\end{wrapfigure}

Section \label{sec:experimentresults} describes the tests conducted to
examine the effect of dictionary, document, and corpus size on the
performance of a single Aho-Corasick indexer and in \label{sec:naive}
we compare our indexer performance to that of a na\"{\i}ve
implementation. Construction time of the state machine as a function
of dictionary size was tested in order to quantify the overhead it
imposes on the indexing task. In section \ref{parallelization} we
explore the performance characteristics of a parallel indexer
deployment.

All tests were run on a single Intel Core 2 Duo 2GHz processor, 1GB
RAM, running a 64-bit Linux 2.6.35 SMP kernel. The Java Virtual
Machine used was version 1.6.0-24.

%----------------------------------------
% Experiment Results
%----------------------------------------
\section{Experiment Results}
\label{sec:experimentresults}

%----------------------------------------
% Number of Keywords
%----------------------------------------
\subsection{Number of Keywords}
To study the impact of dictionary size on performance keywords, a
10,000 keyword dictionary of the most popular English words
\cite{wordlist} and a corpus of 50,000 blog posts of varying 
size were used. First, the time taken to construct the Aho-Corasick state
machine was measured for different numbers of dictionary keywords. The
results of this test, shown in figure
\ref{fig:numkeywordstimecomplexbuildsm}, indicate construction time to be
linear with respect to the number of keywords. This is consistent with
\cite{RefWorks:103} which formally proves that the state machine
construction algorithm is linearly proportional to the sum of the
lengths of the keywords used to construct the state machine.

@@

Second, the time taken for a state machines to build an inverted index
of our corpus was measured. The test was repeated using state machines
constructed with a varying number of keywords and the results plotted
in figure \ref{fig:numkeywordstimecomplexbuildindex}. According to
\cite{RefWorks:103}, which proves that the number of state transitions
involved in processing an input string is independent of the number of
keywords used to construct the state machine, we would not expect the
time complexity to increase with the size of the keyword set. However,
as can be seen in \ref{fig:numkeywordstimecomplexbuildindex}, the time
to build the index increases logarithmically with the number of
keywords. This is likely due to the fact that as the number of
keywords increases, so does the size of the index and the amount of
database interactions required to update the index.

It is important to note that for this experiment, the time taken to
construct the state machine is two orders of magnitude lower than the
time taken by the indexing task. This is an important result as it
validates the practical use of the Aho-Corasick algorithm to index
large corpora, the overhead of state machine construction time being
negligible relative to the indexing task. Also, the space used by the
index in the database is proportional to the number of words we are
using.

%----------------------------------------
% Size of Corpus
%----------------------------------------
\subsection{Size of Corpus}

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.70\textwidth,height=!]{corpsizetimecomplexbuildindex}
  \end{center}
    \caption{Time to build the inverted index as a function of the
      size of the corpus. Time taken is shown in milliseconds on the
      y-axis and the number of documents in the collection on the
      x-axis} 
    \label{fig:corpsizetimecomplexbuildindex}
\end{wrapfigure}

To examine the effect of corpus size on the performance of the
indexer, a 5,000 keyword dictionary and a corpus of 800,000 blog posts
of various sizes were used. The timing results of indexing corpora of
varying sizes, shown in figure
\ref{fig:corpsizetimecomplexbuildindex}, indicate that the time to
build the index with the state machine is linear with respect to the
number of posts.

%----------------------------------------
% Document Size
%----------------------------------------
\subsection{Document Size} 
To examine the effect of document size on indexer performance, 
a 5,000 keyword dictionary and 5,000 blog posts were used. Tests were
run in which the the size of the blog posts was varied, generating the
performance curve in figure \ref{fig:docsizetimecomplexbuildindex}.  
As expected the time to build the indexer is almost linear with
respect to the average document size.

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{docsizetimecomplexbuildindex}
    \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis} 
    \label{fig:docsizetimecomplexbuildindex}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{docsizespacecomplexbuildindex}
    \caption{Size of index as a function of the average document
      size. The number of database rows used to store the index is
      shown on the y-axis and the average document size in characters
      on the x-axis}  
    \label{fig:docsizespacecomplexbuildindex}
  \end{minipage}
\end{figure}

Examining the size of the index created, the larger the documents the
more records created. However, the relationship is not linear since a
single record is stored to represent one or many occurrences of a
particular keyword in a document. This relationship is shown in figure  
\ref{fig:docsizespacecomplexbuildindex}.

%----------------------------------------
% Comparison with Naive Index Builder
%----------------------------------------
\subsection{Comparison with Na\"{i}ve Index Builder}
\label{sec:naive}
In order to quantify the benefits of using the Aho-Corasick algorithm
for indexing, our indexer performace was compared to that of a
na\"{i}ve implementation. The na\"{i}ve indexer builder has the same
input as our system i.e. a list of keywords and a list of documents,
and the same output i.e. an inverted index corresponding to the
keywords and documents. Algorithm \ref{alg:invertedindex} is used in
principal for both implementations. However, whereas the Aho-Corasick
implementation uses a state machine to scan, the na\"{\i}ve
indexer tokenizes the document and compares each token to the keyword
dictionary.

@@

Algorithm \ref{alg:invertedindex} has a complexity proportional to the
number of keywords $n$, the number of document $m$, and the average
document size $\bar{c}$. For our chosen architecture, execution time
is also effected by database latency. In the case of the
na\"{\i}ve indexer, we perform one database transaction per keyword
and in the Aho-Corasick implementation, we perform a database
transaction per document. For analysis purposes, we designate the
average database latency per transaction as $\bar{t}$.

For the na\"{i}ve indexer implementation described above, the expected
time complexity relationship is shown in equation
\ref{eqn:timecomplexworstcase}. For the Aho-Corasick indexer, the
expected theoretical complexity relationship is given by equation
\ref{eqn:timecomplexsm}.

\begin{equation}
\label{eqn:timecomplexworstcase}
Time \propto n * ( m * \bar{c} + \bar{t} )
\end{equation}

\begin{equation}
\label{eqn:timecomplexsm}
Time \propto m * ( \bar{c}  + \bar{t} )
\end{equation}

To examine the effect of document size on idexer performance, the number of keywords was
fixed at 1,000, the number of documents at 5,000, and tests run
against corpora composed of varying average length. From equations
\ref{eqn:timecomplexworstcase} and \ref{eqn:timecomplexsm}, we would
expect both implementations to behave linearly with repect to average
documetn size and to differ by constant scaling factor $n$, the size
of the keyword dictionary. The experiment results are shown in figure
\ref{fig:naivedocumentsize} and support this analysis. Also, we can observe
that for smaller documents, the na\"{\i}ve method
performs better. This is attributable to the fact that for smaller
documents, the database advantage of the na\"{\i}ve implementation is
enough to overcome the keyword dictionary size scaling factor. To see
this effect more clearly, experiment results are plotted on a
logarithmic scale in figure \ref{fig:naivedocumentsizelog}.

@@

xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

@@

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.60\textwidth,height=!]{parallelization}
  \end{center}
  \caption{Performance characteristics of parallel indexer
    deployment. The \textit{ideal} curve shows the theoretical
    performance curve based on linear combination of average
    individual node performance over the corpus} 
  \label{fig:parallelization}
\end{wrapfigure}
